Structural and Geographic Influences on Residential Parking Availability in Toronto

### 2024-12-01 paper/paper.qmd
added consideration of alternative models and model justification

Residential parking availability is a critical aspect of urban planning, particularly in dense metropolitan areas like Toronto. The rise in vehicle ownership and the increasing demand for parking spaces have created challenges for both homeowners and city planners. This paper examines the factors influencing the availability of more than one parking space in residential properties, using a dataset from Toronto’s Open Data Portal that documents front yard parking permits and related attributes. While parking availability has been explored in terms of policy and infrastructure, few studies have systematically analyzed how structural and geographic factors combine to affect the likelihood of multiple parking spaces at a property. This paper addresses this gap by applying a logistic regression framework to examine how location and parking type impact this outcome.

@@

This model is well-suited to the research question because the outcome variable, whether a property has more than one parking space, is binary. A logistic regression framework naturally models the relationship between this binary outcome and the predictors. Parking type and geographic location are included as predictors because they capture structural and locational influences on parking availability. Weakly informative priors ensure that the model is not overly constrained by prior beliefs, allowing the observed data to guide the estimates while guarding against overfitting.

@@

first_model <-
  readRDS(file = here::here("models/logistic_model.rds"))
```

@@

modelsummary::modelsummary(
  list(
    "Logistic regression model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

This paper provides a detailed examination of the factors that influence whether a residential property in Toronto has more than one parking space. By employing a Bayesian logistic regression model, we uncover critical insights into how parking type and geographic location shape parking availability. One key finding is that structural factors, such as parking type, exert a strong influence on whether properties can accommodate multiple parking spaces. Specifically, properties with boulevard parking are more likely to have additional spaces compared to those with front yard or widened driveway parking, which tend to be more spatially constrained. This emphasizes the importance of physical characteristics in determining parking capacity and provides practical implications for urban planning and residential design.

@@

Another important finding relates to the role of geographic location, captured through ward-level data. While the likelihood of a property having more than one parking space is generally low across all wards, certain areas, such as York South-Weston and Etobicoke Centre, exhibit slightly higher probabilities. These differences may reflect variations in zoning regulations, housing density, and property size across Toronto’s diverse neighborhoods. By highlighting these geographic disparities, this paper contributes to our understanding of how local policies and urban layouts affect parking availability. These insights could inform targeted zoning reforms and development strategies in areas where parking demand outstrips supply.

@@

Beyond the specific findings, this analysis sheds light on broader urban planning challenges associated with residential parking. The increasing demand for parking in growing metropolitan areas like Toronto underscores the need for innovative solutions that balance parking availability with land use efficiency. Our results suggest that policymakers should consider both structural and locational factors when addressing parking shortages. For example, incentivizing the adoption of space-efficient parking designs, such as boulevard parking, or revising zoning policies to allow for greater parking capacity in underserved wards could help address these challenges. This study demonstrates how data-driven approaches can guide such policy decisions, offering a framework for similar analyses in other cities.

@@

While this study provides valuable insights, it is not without limitations. First, the dataset relies on administrative records, which may not capture informal or unlicensed parking arrangements. This could lead to an underestimation of the true availability of additional parking spaces. Second, the analysis focuses on a limited set of predictors, leaving out potentially relevant factors such as property size, household income, or neighborhood-level car ownership rates. Incorporating these variables in future research could enhance the explanatory power of the model. Finally, the geographic scale of analysis is restricted to wards, which may overlook finer-grained spatial patterns within neighborhoods. Future studies could explore these patterns at a more localized level, providing a deeper understanding of parking dynamics and informing more targeted interventions. These next steps would build on the findings of this paper, advancing our understanding of residential parking and its implications for urban policy.

>>>>

Residential parking availability plays a pivotal role in shaping urban mobility and land use policies, particularly in dense cities like Toronto. Previous studies have highlighted the importance of residential parking permits and supply in influencing local traffic conditions and land allocation [@van2014residential]. For instance, the convenience of home parking has been shown to significantly impact household car usage, underscoring the interplay between parking infrastructure and car dependency [@guo2013home]. Moreover, research has demonstrated that vibrant urban districts often face unique challenges in balancing residential parking needs with limited space, requiring tailored policies to accommodate both residents and visitors [@molenda2013residential]. Building on this body of work, this study investigates how structural and locational factors affect the likelihood of residential properties having more than one parking space, providing insights for more efficient and equitable urban planning. Residential parking availability is a critical aspect of urban planning, particularly in dense metropolitan areas like Toronto. The rise in vehicle ownership and the increasing demand for parking spaces have created challenges for both homeowners and city planners. This paper examines the factors influencing the availability of more than one parking space in residential properties, using a dataset from Toronto’s Open Data Portal that documents front yard parking permits and related attributes. While parking availability has been explored in terms of policy and infrastructure, few studies have systematically analyzed how structural and geographic factors combine to affect the likelihood of multiple parking spaces at a property. This paper addresses this gap by applying a logistic regression framework to examine how location and parking type impact this outcome.

@@

This model is well-suited to the research question because the outcome variable, whether a property has more than one parking space, is binary. A logistic regression framework naturally models the relationship between this binary outcome and the predictors, capturing the log-odds of the outcome as a linear combination of the predictors. The Bayesian implementation of logistic regression adds further advantages, allowing for the inclusion of prior distributions to reflect uncertainty or prior knowledge about the relationships between predictors and the outcome. Weakly informative priors, as used in this model, ensure that the analysis remains data-driven while providing regularization to prevent overfitting. Additionally, the Bayesian framework yields posterior distributions for each parameter, offering richer insights into the uncertainty and variability of the estimates, which is particularly valuable in contexts like urban planning where decisions often involve uncertainty.

The inclusion of parking type and geographic location as predictors is highly appropriate given their theoretical and practical relevance to the research question. Parking type reflects structural constraints, as different designs (e.g., boulevard parking versus front yard parking) inherently allow for varying capacities. Geographic location, represented by ward, captures locational influences, including differences in zoning regulations, housing density, and property sizes across Toronto. Together, these predictors account for both the physical design of properties and the broader spatial context in which they exist, ensuring a comprehensive analysis of parking availability. Furthermore, these variables align with established findings in urban planning literature, which emphasize the role of structural and geographic factors in shaping parking dynamics. By incorporating these predictors, the model not only provides actionable insights but also positions the findings within a robust theoretical framework, enhancing their relevance for policy and planning.

### Comparison with alternative models

The process involved splitting the dataset into training and testing sets, with 80% of the data used for training and 20% for testing. Both Bayesian and Frequentist logistic regression models were trained using the training set, with the Bayesian model incorporating priors and the Frequentist model relying solely on observed data. To evaluate prediction accuracy, the models were applied to the testing set to predict whether a property had more than one parking space. Probabilities were converted into binary predictions using a threshold of 0.5, and accuracy was calculated as the proportion of correct predictions. The results showed identical accuracy for both models at approximately 95.52%, suggesting that both approaches performed equally well in classifying the test data. This parity indicates that the Bayesian model’s priors and the Frequentist model’s reliance on maximum likelihood produced comparable predictive capabilities for this dataset.

While the Bayesian and Frequentist logistic regression models demonstrated identical accuracy in this analysis, there are compelling reasons to prefer the Bayesian approach in this context. The Bayesian model allows for the incorporation of prior information, which can be particularly advantageous when sample sizes are smaller or when there is prior domain knowledge about the effects of variables, such as parking type or ward characteristics. Even when priors are set to be non-informative, as in this case, the Bayesian framework naturally accounts for uncertainty in parameter estimates, offering a richer interpretation through posterior distributions rather than point estimates alone. This additional layer of insight is particularly valuable in urban planning contexts, where decisions often involve uncertainty and the need to weigh competing priorities.

Furthermore, Bayesian models provide posterior predictive distributions, allowing for a nuanced understanding of the probability of outcomes, rather than relying solely on a single predicted value. This can be especially useful for stakeholders seeking to quantify risks or probabilities associated with certain scenarios, such as the likelihood of a property having more than one parking space in specific wards. The flexibility of Bayesian modeling to incorporate future data or adjust priors also ensures that it remains robust and adaptive as more information becomes available, making it an ideal choice for iterative and long-term urban studies. While the identical accuracy suggests comparable predictive performance, the Bayesian model’s ability to quantify uncertainty and provide a deeper probabilistic understanding justifies its use as the preferred approach.

@@

Bayesian_logit_model <-
  readRDS(file = here::here("models/Bayesian_logistic_model.rds"))
```

@@

modelsummary::modelsummary(
  list(
    "Bayesian logistic regression model" = Bayesian_logit_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

This study undertakes a comprehensive examination of the factors influencing the availability of more than one parking space for residential properties in Toronto, offering new insights into the intricate interplay of structural design and geographic location. Employing a Bayesian logistic regression framework, the analysis highlights the pivotal role of parking type as a determinant of parking capacity. Properties featuring boulevard parking demonstrate a markedly higher likelihood of accommodating multiple spaces, a result that underscores the inherent spatial flexibility provided by this design. In contrast, front yard parking and widened driveways are constrained by their physical configurations, limiting their potential for expansion. These findings reflect broader implications for residential design, suggesting that spatial planning at the individual property level can have a profound impact on the ability of urban housing to meet growing parking demands. Such insights are particularly relevant for urban planners, architects, and policymakers aiming to optimize land use while addressing the practicalities of residential parking.

Additionally, this analysis sheds light on the indirect implications of structural decisions, which often intersect with broader urban design principles. The strong influence of parking type highlights the need for a more deliberate approach to residential development that integrates parking considerations into zoning and building guidelines. For instance, prioritizing the adoption of boulevard parking in new residential developments could alleviate some of the pressures associated with parking shortages, particularly in high-demand areas. These results also call attention to the importance of aligning residential parking policies with urban growth strategies, ensuring that structural characteristics of properties are leveraged to address both current and future parking challenges.

@@

The geographic component of this analysis further reveals important variations in parking availability, demonstrating how broader spatial and regulatory contexts shape residential parking dynamics. By leveraging ward-level data, the study identifies nuanced differences in the probability of properties offering multiple parking spaces across Toronto’s diverse neighborhoods. Notably, wards such as York South-Weston and Etobicoke Centre exhibit slightly elevated probabilities of additional parking capacity. These geographic disparities likely reflect a confluence of factors, including zoning regulations, housing density, and property size, as well as historical urban development patterns. For example, suburban wards may benefit from larger lot sizes and more lenient zoning requirements, enabling higher parking capacity, whereas more densely populated areas face physical and regulatory constraints that limit parking expansion.

This spatial variability underscores the importance of tailoring policy interventions to the unique characteristics of each neighborhood. Localized reforms, such as revisiting zoning codes or incentivizing parking capacity enhancements in areas with acute shortages, could mitigate the uneven distribution of parking resources. Additionally, this analysis highlights the broader urban planning challenge of reconciling the tension between accommodating private vehicles and promoting compact, sustainable urban design. By focusing on the geographic nuances of parking availability, policymakers can develop more targeted and contextually appropriate strategies that align with both local needs and broader sustainability goals.

@@

Beyond the specific structural and geographic insights, this study contributes to a broader understanding of the complex challenges surrounding residential parking in growing metropolitan areas like Toronto. The rapid increase in vehicle ownership, coupled with the finite availability of urban space, necessitates innovative solutions that balance the competing demands of accessibility and efficiency. This study’s findings point to the critical need for integrated approaches that account for both structural and locational factors in addressing parking shortages. For example, incentivizing developers to adopt space-efficient designs, such as boulevard parking, could not only increase parking availability but also make more effective use of limited urban land. Similarly, revising zoning policies to allow for greater parking capacity in underserved neighborhoods could help alleviate geographic inequities in parking access.

The findings also underscore the value of data-driven decision-making in urban planning. By systematically analyzing how structural and geographic factors shape parking capacity, this study offers a replicable framework for evaluating residential parking challenges in other urban contexts. Moreover, the incorporation of advanced modeling techniques, such as Bayesian logistic regression, provides a robust foundation for exploring the multifaceted dynamics of parking availability. Policymakers and urban planners can draw on these insights to craft interventions that are not only effective in addressing immediate challenges but also aligned with long-term urban sustainability objectives.

@@

While the findings of this study offer valuable contributions to understanding residential parking dynamics, certain limitations warrant careful consideration. One key limitation is the reliance on administrative records, which may fail to capture unlicensed or informal parking arrangements. Such omissions could result in an underestimation of parking availability, particularly in areas where informal practices, such as shared driveways or unauthorized extensions, are prevalent. Future research could address this gap by incorporating alternative data sources, such as satellite imagery or field surveys, to provide a more comprehensive view of parking conditions.

Another limitation is the relatively narrow scope of predictors used in the analysis. While parking type and geographic location are critical factors, additional variables, such as property size, household income, or neighborhood-level car ownership rates, could offer deeper insights into the determinants of parking availability. Expanding the model to include these variables would not only enhance its explanatory power but also provide a more nuanced understanding of the socioeconomic and spatial dynamics of residential parking. Furthermore, the geographic resolution of this study, which focuses on ward-level analysis, may overlook important variations within smaller spatial units. Future research could benefit from a finer-grained approach, examining parking patterns at the neighborhood or even street level to uncover localized trends that are masked at broader scales.

Building on these limitations, future studies could also explore the interplay between residential parking and environmental factors, such as public transit accessibility or proximity to commercial hubs. Integrating qualitative data, such as surveys capturing residents’ attitudes and preferences regarding parking, could complement quantitative findings and provide a richer understanding of the underlying drivers of parking behavior. These next steps would build on the insights of this study, advancing the development of targeted, equitable, and sustainable strategies for managing residential parking in rapidly urbanizing cities.
